---
title: "Disaster Relief Simulation"
author: "Daniel Dittenhafer & Jashan Narhan"
date: "April 26, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 604: Simulation & Modeling"
---
```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R"), results='hide',message=FALSE}
# Initialize our Rmd environment
```

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(RefManageR)

cleanbib()

cite_options(style="markdown")

# bibBirthData <- bibentry(bibtype="Misc",
#                          author=person(family="HHS"),
#                          publisher="United States Department of Health and Human Services (US DHHS), Centers for Disease Control and Prevention (CDC), National Center for Health Statistics (NCHS), Division of Vital Statistics",
#                          title="Natality public-use data 2007-2014 on CDC WONDER Online Database",
#                          year=2016,
#                          month="February",
#                          url="http://wonder.cdc.gov/natality-current.html")


bibBastianEtAl <- bibentry(bibtype="Article",
                 title="Multi-criteria logistics modeling for military humanitarian assistance and disaster relief aerial delivery operations",
                 author=personList(person(family="Bastian", given="Nathaniel"),
                                   person(family="Griffin", given="Paul"),
                                   person(family="Spero", given="Eric"),
                                   person(family="Fulton", given="Lawrence")),
                 publisher="Springer Berlin Heidelberg",
                 journal="Optimization Letters",
                 year=2015,
                 doi="10.1007/s11590-015-0888-1")

bibOzdamarEtAl <- bibentry(bibtype="Article",
                 title="Emergency Logistics Planning in Natural Disasters",
                 author=personList(person(family="Ozdamar", given="Linet"),
                                   person(family="Ekinci", given="Ediz"),
                                   person(family="Kucukyazici", given="Beste")),
                 publisher="Kluwer Academic Publishers",
                 journal="Annals of Operations Research",
                 year=2004)

bibHdxWeb <- bibentry(bibtype="Misc",
                      author=person(family="United Nations Secretariat"),
                      title="Humanitarian Data Exchange",
                      publisher="United Nations Office for the Coordination of Humanitarian Affairs",
                      year=2016,
                      url="https://data.hdx.rwlabs.org/")

bibCREDWeb <- bibentry(bibtype="Misc",
                      author=person(family="EM-Dat"),
                      title="International Disaster Database",
                      publisher="Centre for Research on the Epidemiology of Disasters",
                      year=2016,
                      url="http://www.emdat.be/advanced_search/index.html")

  
```

```{r helper-funcs, echo=FALSE}
reliefBenAvgForScenarios <- function(scenarios)
{
  dfExp <- subset(dfResults, Object.Name == "ReliefBeneficiary" & 
                            Scenario %in% scenarios &
                            Data.Item == "TimeInSystem" &
                            Statistic.Type == "Average (Hours)")  
  return(dfExp)
}
```

# Abstract

# Keywords

# Literature Review

The number of disasters witnessed over time appear to be increasing in terms of 
frequency and impact. Figure 1 for example, depicts estimates of the average 
number of people impacted (i.e. made homeless, injured or economically affected)
or killed from natural disaster from the year 1900 to date, based on disaster 
data pertaining to notable events in each year `r citep(bibCREDWeb)`.

\begin{center}
\includegraphics{../images/Report_Fig1.png}
\end{center}
\begin{center}
\text{Fig 1: Averaged number of people killed or affected by disasters by year}
\end{center}

While there are fluctuations from year to year, there is a clear trend showing a
growth in the numbers or people who are affected or killed by disasters over 
time. 

Responding to such disasters is a highly complex process. Often multiple 
stakeholders will be engaged, working under stressful environments with severe 
time pressure and significant uncertainty with respect to whether they can work
to save lives. In this context, we are beginning to see an increase in the 
amount of research into models and systems that can potentially help aid workers
in making informed decisions.

... summarize the research briefly TBC.

# Methodology

## Structure

### Supplier

A Supplier entity is included in the simulation model. This entity is the source of ReliefSupplies. This entity is used, through design of experiments, to simulate both reliable partner suppliers as well as ad hoc suppliers (more variability in their supply). The following variables (Referenced Properties) were created to facilitate the study: 

* The time to deliver a shipment of supplies:  *SupplierTravelTimeToWarehouse*. 

* The maximum number of relief supplies which the supplier will provide: *SupplierMaximumArrivals* 

The delivery vehicle, PartnerVehicle, was set to require a full load prior to departing for the Warehouse.

* The capacity of the PartnerVehicle: *SupplierVehicleCapacity*

### Warehouse

A Warehouse was added to store disaster relief supplies in preparation for a disaster. This entity was modelled as a server which processes incoming ReliefSupplies and stores them in a ready state for movement to a DistributionSite.

The following variables were created to facilitate design of experiments:

* The storage capacity of the warehouse: *WarehouseStorageCapacity*

The delivery vehicle, WarehouseVehicle, was set to require a full load prior to departing for a distribution site and have periodic failures.

* The capacity of the WarehouseVehicle: *WarehouseVehicleCapacity*

* The frequency of failure: *WarehouseVehicleUptimeBetweenFailures*

The selection weight on the paths from the Warehouse to the Distribution Sites were set to dynamically adjust based on the number of Relief Beneficiaries  waiting for Relief Supplies.

### Distribution Sites

Two distribution sites are included in the simulation and are modelled as combiners which match Relief Supplies to Relief Beneficiaries.

The following variables were created to facilitate design of experiments:

* The processing time needed to provide a ReliefSupply item to a ReliefBeneficiary: DistSiteProcessingTime

### Relief Beneficiaries

The Relief Beneficiaries represent entities who are affected by the disaster and require ReliefSupplies. Two separate sources were included in the model to simulate separate points from which beneficiaries would origination for each distribution site.

The following variables were created to facilitate design of experiments:

* The interarrival time of the ReliefBeneficiaries: *ReliefBeneficiaryInterarrivalTime*

* The maximum number of ReliefBeneficiaries from a given source: *DisasterVictimMaximumArrivals*

### Disaster

The "Disaster" is set to begin at a designated time into the simulation. This gives some time to prestage relief supplies through the warehouse (priming the pump).

The following variables were created to facilitate design of experiments:

* The amount of time between simulation start and onset of the disaster: *DisasterTimeOffset*


## Measured Outcomes

### Average Beneficiary Time in System

Measured in hours, this response is an important indicator of the performance of the system.

### Relief Supplies Distributed

How many relief supplies were distributed?

### Unsatisfied Beneficiaries

Naturally, we wanted this response to be zero (0) throughout all simulations.

### Maximum Distribution Site 1 Time Waiting

Measured in hours, this response helped us understand the degree of the worst case scenario for distribution site 1 beneficiaries in need of relief supplies.

### Maximum Distribution Site 2 Time Waiting

Measured in hours, this response helped us understand the degree of the worst case scenario for distribution site 2 beneficiaries in need of relief supplies.

## Design of Experiments

### Supplier: Partner vs Ad Hoc

How does a dedicated supplier whose production time is lower (more units of production are committed to our orders), versus an ad hoc supplier whose production time is longer, affect the measured outcomes.

### Control Variables

The following table shows key parameters (control variables) for each scenario:

```{r, echo=FALSE}
dfControls <- data.frame(
  Name=c("Partner Supplier", 
         "Adhoc Supplier", 
         "Midlevel Supplier", 
         "Midlevel Access to Dist Site 1", 
         "Quick Access to Dist Site 1"),
  ReliefSuppliesInterarrivalTime=c("Random.Normal(1, 0.1)",
                                   "Random.Normal(4, 0.1)",
                                   "Random.Normal(2.5, 0.1)",
                                   "Random.Normal(1, 0.1)",
                                   "Random.Normal(1, 0.1)"),  WarehouseTravelTimeToDistSite1=c("Random.Normal(6,0.5)",
                                   "Random.Normal(6,0.5)",
                                   "Random.Normal(6,0.5)",
                                   "Random.Normal(4,0.5)",
                                   "Random.Normal(2,0.5)")

)
kable(dfControls)
```

# Results

```{r, echo=FALSE}
# Load the results from the Simio CSV
dfResults <- read.table("../project/DisasterReliefSim-Simio_DisasterReliefModel_SupplierExperiment_ResultsSummary.csv",
                        sep=",", header=TRUE, stringsAsFactors=FALSE)

#summary(dfResults)

# Add on our Control Variable values (haven't found how to export these from Simio yet)
dfResults$SupplierReliefSuppliesInterarrivalTimeMean <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 1
dfResults[dfResults$Scenario == "Midlevel Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 2.5
dfResults[dfResults$Scenario == "Adhoc Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 4

dfResults$WarehouseTravelTimeToDistributionSite1 <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$WarehouseTravelTimeToDistributionSite1 <- 6
dfResults[dfResults$Scenario == "Midlevel Access to Dist Site 1", ]$WarehouseTravelTimeToDistributionSite1 <- 4
dfResults[dfResults$Scenario == "Quick Access to Dist Site 1", ]$WarehouseTravelTimeToDistributionSite1 <- 2
```

## Supplier Production Committment

```{r, echo=FALSE, fig.height=3}
# Select only the scenarios that are related to this experiment
vSupplierScenarios <- c("Partner Supplier", "Adhoc Supplier", "Midlevel Supplier")
dfExp <- reliefBenAvgForScenarios(vSupplierScenarios)

g1 <- ggplot(dfExp) + 
  geom_line(aes(x=SupplierReliefSuppliesInterarrivalTimeMean, y=Average)) +
  labs(x="Supplier Relief Supplies Interarrival Time Mean (minutes)",
       y="Avg Beneficiary Time In System (hours)") +
  myTheme
g1
```

## Warehouse Access to Distribution Site

```{r, echo=FALSE, fig.height=3}
vScenarios <- c("Partner Supplier", 
                "Midlevel Access to Dist Site 1", 
                "Quick Access to Dist Site 1")

# Select only the scenarios that are related to this experiment
dfExp <- reliefBenAvgForScenarios(vScenarios)

#dfExp
g2 <- ggplot(dfExp) + 
  geom_line(aes(x=WarehouseTravelTimeToDistributionSite1, y=Average)) +
  labs(x="Supplier Relief Supplies Interarrival Time Mean (minutes)",
       y="Avg Beneficiary Time In System (hours)") +
  myTheme
g2
```

# Summary

# References

```{r, results='asis', echo=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

# Appendix: Raw Code

