---
title: "Intracountry Disaster Relief Simulation"
author: "Daniel Dittenhafer & Jashan Narhan"
date: "May 22, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 604: Simulation & Modeling"
---
```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R"), results='hide',message=FALSE, warning=FALSE}
# Initialize our Rmd environment

```

```{r, echo=FALSE, message=FALSE}

bibBastianEtAl <- bibentry(bibtype="Article",
                 title="Multi-criteria logistics modeling for military humanitarian assistance and disaster relief aerial delivery operations",
                 author=personList(person(family="Bastian", given="Nathaniel"),
                                   person(family="Griffin", given="Paul"),
                                   person(family="Spero", given="Eric"),
                                   person(family="Fulton", given="Lawrence")),
                 publisher="Springer Berlin Heidelberg",
                 journal="Optimization Letters",
                 year=2015,
                 doi="10.1007/s11590-015-0888-1")

bibOzdamarEtAl <- bibentry(bibtype="Article",
                 title="Emergency Logistics Planning in Natural Disasters",
                 author=personList(person(family="Ozdamar", given="Linet"),
                                   person(family="Ekinci", given="Ediz"),
                                   person(family="Kucukyazici", given="Beste")),
                 publisher="Kluwer Academic Publishers",
                 journal="Annals of Operations Research",
                 year=2004,
                 doi="10.1023/B:ANOR.0000030690.27939.39")

bibKungEtAl <- bibentry(bibtype="Article",
                 title="Designing intelligent disaster prediction models and systems for debris-flow in Taiwan",
                 author=personList(person(family="Kung", given="Hsu-Yang"),
                                   person(family="Chen", given="Chi-Hua"),
                                   person(family="Ku", given="Hao-Hsiang")),
                 publisher="Elsevier",
                 journal="Expert Systems with Applications",
                 year=2012, 
                 doi="10.1016/j.eswa.2011.11.083")


bibRottEtAl <- bibentry(bibtype="Article",
                 title="A transshipment model for distribution and inventory relocation under
                 uncertainty in humanitarian operations",
                 author=personList(person(family="Rottkemper", given="Beate"),
                                   person(family="Fisher", given="Kathrin"),
                                   person(family="Blecken", given="Alexander")),
                 publisher="Elsevier",
                 journal="Socio-Economic Planning Sciences",
                 year=2012,
                 doi="10.1016/j.seps.2011.09.003")

bibHdxWeb <- bibentry(bibtype="Misc",
                      author=person(family="United Nations Secretariat"),
                      title="Humanitarian Data Exchange",
                      publisher="United Nations Office for the Coordination of Humanitarian Affairs",
                      year=2016,
                      url="https://data.hdx.rwlabs.org/")

bibCREDWeb <- bibentry(bibtype="Misc",
                      author=person(family="EM-Dat"),
                      title="International Disaster Database",
                      publisher="Centre for Research on the Epidemiology of Disasters",
                      year=2016,
                      url="http://www.emdat.be/advanced_search/index.html")

bibGitHub <- bibentry(bibtype="Misc",
                      title="DisasterReliefSim",
                      author=personList(person(family="Dittenhafer", given="Daniel W"), 
                                   person(family="Narhan", given="Jashan")),
                      year=2016,
                      url="https://github.com/dwdii/DisasterReliefSim")
  
```

```{r helper-funcs, echo=FALSE}
reliefBenAvgForScenarios <- function(scenarios, data)
{
  dfExp <- subset(data, Object.Name == "ReliefBeneficiary" & 
                            Scenario %in% scenarios &
                            Data.Item == "TimeInSystem" &
                            Statistic.Type == "Average (Hours)")  
  return(dfExp)
}

supplyAvgForScenarios <- function(scenarios, data)
{
  dfExp <- subset(data, Object.Name == "ReliefSupplyItem" & 
                            Scenario %in% scenarios &
                            Data.Item == "TimeInSystem" &
                            Statistic.Type == "Average (Hours)")  
  return(dfExp)
}

dataForScenarios <- function(scenarios, data)
{
  dfData <- subset(data,  Scenario %in% scenarios )
  dfData <- dfData[,c("Scenario", 
                      "Response", 
                      "Mean", 
                      "Mean.Confidence.Interval.Start",
                      "Mean.Confidence.Interval.End")]
  colnames(dfData) <- c("Scenario", 
                      "Response", 
                      "Mean", 
                      "95% CI Lwr",
                      "95% CI Upr")
  rownames(dfData) <- NULL
  return(dfData)
}
```

# Abstract

As disasters have affected more people, research into various aspects of relief planning and delivery appear to be increasing. This study focuses on in-country supply chain logistics extending from the originating supplier to last mile distribution to beneficiaries. A simulation model is developed using Simio and experiments are performed to gauge sensitivity to various parameters and decision variables. Supplier relationship factors are found to be key drivers of beneficiary outcomes. To facilitate reproducible research, the research materials are shared on GitHub.

# Keywords

humanitarian aid, logistics, disaster relief, simulation

# Literature Review

The number of disasters witnessed over time appear to be increasing in terms of 
frequency and impact. Figure 1 for example, depicts the average number of people
killed or adversely impacted (i.e. made homeless, injured or economically 
affected) due to natural disasters from the year 1900 to date. This information 
is based on disaster data pertaining to *notable* events in each year 
`r citep(bibCREDWeb)`.

```{r, echo=FALSE, fig.height=3}
dat <- read.csv('../data/DataByYear.csv')
dat <- dat[, -c(2,9)]

dat[is.na(dat)] <- 0  ## NAs are being replaced as zeros for the purpose of 
                      ## figure generation
dat$Total <- rowSums(dat[,3:6])  ## Aggegating killed and affected (made 
                                 ## homeless, affected, injured)
dat <- dat[, c(1, 8)]

df <- dat %>% group_by(year) %>% summarise(count = n(), KillAffect = sum(Total))
df$Avg.KillAffected <- (df$KillAffect/df$count)/1e+6

# generate the plot
g0 <- ggplot(df, aes(x = year, y = Avg.KillAffected)) + geom_line() + 
        ylab("People (Millions)") + xlab("") +
        ggtitle("Figure 1: Average Numbers Killed or Affected By Major Disasters 
                in Given Year") +
        myTheme
g0
```

While there are fluctuations from year to year, there is a clear trend showing a
growth in the numbers or people who are affected or killed by disasters over 
time. 

Responding to these types of disasters is a highly complex process. Often 
multiple stakeholders will be engaged, working under stressful environments with
severe time pressure and significant uncertainty with respect to whether they 
can work to save lives. 

In this context, we are beginning to see an increase in the amount of research 
into models and systems that can potentially help humanitarian workers make 
informed decisions. For example Kung et al `r citep(bibKungEtAl)` present three 
prediction models and an inference engine using linear regression, multivariate 
analysis and back propagation networks to assess potential debris flows 
resulting from earthquakes. In another example, Rottkemper et al 
`r citep(bibRottEtAl)` present a mixed-integer programming model for an 
integrated relocation and distribution planning solution designed to  minimize 
both operational costs and unmet demand for relief items.

Within the *logistics* aspects of emergency management, Ozdamar et al 
`r citep(bibOzdamarEtAl)` present a model to complement decision support systems
related to logistics planning following a disaster. The study contrasts various 
algorithmic approaches to solving for multi-period, multi-commodity network flow
problems coupled with a vehicle routing problem.

Ozdamar et al's study helps to designate routing decisions for vehicles. The 
authors also note that the resolution to the various optimization choices are 
NP-hard. They suggest heuristic methodologies be applied to find solutions to 
large scale problems. Building on this recommendation, the current study, aims 
to understand the dynamics of in-country humanitarian logistics problems via 
simulations.

In running the simulation, we present a model of a simplified humanitarian 
logistics chain for the purpose of understanding the behaviour of the system and 
for evaluating certain strategies. We modelled key aspects of the 
humanitarian logistics chain using computer simulation software (Simio v8.136.13435). The 
intent is to understand the realities of moving humanitarian aid, particularly 
to the last mile of distributions (something that has not garnered as much 
research in the literature). 

# Methodology

## Structure

The basic objective of a logistics system is to deliver the appropriate 
supplies, in good condition, in the quantities required, and at the places and 
time they are needed. Some emergencies generate a limited need for very rapid 
and very specific deliveries of supplies, commodities and resources from outside
the affected area (i.e. outside the affected country). This process entails the 
implementation of an international humanitarian logistics chain (IHLC) as 
depicted in the following chart.

\begin{center}
\includegraphics{../images/International_Humanitarian_Log_Chain.jpeg}
\end{center}
\begin{center}
\text{Fig 2: The International Humanitarian Logistics Chain}
\end{center}

The IHLC is an involved process adding significant complexity to the movement of
relief resources - see ``r citep(bibBastianEtAl)` for an example of modelling 
an international logistics chain.

While international humanitarian logistics is a complex process, it is also 
important to note that the majority of relief logistic operations never actually
receive international attention. With this reality in mind, and in an effort to 
simplify the modelling of the logistic chain, this project will consider the 
simulation needs of a in-country based logistic chain. The simplification of the
model is shown in Figure 3.

\begin{center}
\includegraphics{../images/Simplified_HLC.jpeg}
\end{center}
\begin{center}
\text{Fig 3: The Simplified Humanitarian Logistics Chain}
\end{center}

Key entities are readily recognized from Figure 3. A short description follows 
on each entity that is modelled.

### Supplier

A Supplier entity is included in the simulation model. This entity is the source
of ReliefSupplies. The Supplier entity is used, through design of experiments, 
to simulate the reliability of suppliers. We do this by modelling three 
categories of supplier: Partner (Preferential Supplier, representing the 
establishment of strategic supply chain partnership agreements with a given 
supplier such as commitments on the frequency of supplies by a given time, 
agreement on price points etc.), an Adhoc Supplier (representing local suppliers
that aid agencies establish at the onset of a disaster but without any formal 
supplier partnership agreement being established), and a Midlevel supplier 
(representing known suppliers ahead of the disaster but who have not entered 
into a strategic agreement with the given aid agency). 

The inter-arrival times for relief supplies produced by these entities differs 
based on the construct of reliability. Note that the inter-arrival times are 
modelled as Random Normal events with mean arrival time of supplies being faster
for Partner Suppliers, followed by Midlevel suppliers and finally Ad-hoc 
suppliers. The normal distribution was selected under an assumption that 
production of supplies would likely be based on known manufacturing processes.
This is noted as a uncorroborated assumption and would require comparisons of 
the simulation results to real-world supplier data under the context of 
emergencies.

It should also be noted that the entity was also modelled on an assumption of
infinite supply under the constraint of the inter-arrival time capability of the 
supplier, and continuous demand driven by the anticipation of a disaster and 
then the disaster itself. This assumption was felt to be appropriate given the 
fact that the simulation is modelling the rapid onset and response to a disaster
wherein estimating affected beneficiary numbers is very hard to achieve up front.
Consequently modelling a continuous supply of relief items, is considered a fair
reflection of the real-world reality of an aid agency establishing a supply chain
for relief items.

The following variables (Referenced Properties) were created to facilitate the 
study: 

* The time to deliver a shipment of supplies:  *SupplierTravelTimeToWarehouse*. 

* The maximum number of relief supplies which the supplier will provide: *SupplierMaximumArrivals* 

* The time between produced relief supplies available for movement to the warehouse: *SupplierReliefSuppliesInterarrivalTime*

The delivery vehicle, SupplierVehicle, was set to require a full load prior to 
departing for the Warehouse.

* The capacity of the SupplierVehicle: *SupplierVehicleCapacity*

### Warehouse

A Warehouse entity was added to store disaster relief supplies in preparation 
for a disaster. This warehouse acted as the central hub for the aid agency to 
supply distribution sites during disasters (i.e. no secondary level warehouses 
were modelled).

The Warehouse entity was modelled as a server which processes incoming 
ReliefSupplies and stores them in a ready state for movement to a 
DistributionSite.

The following variables were created to facilitate design of experiments:

* The storage capacity of the warehouse: *WarehouseStorageCapacity*

The delivery vehicle, WarehouseVehicle, was set to require a full load prior to 
departing for a distribution site. A capacity of 1000 units was defined on these
vehicles. Given the need to control against the risk of transportation being 
hijacked while on route (losing supplies) and the fact that these vehicles need
to access smaller roads, the warehouse to distribution sites are modelled with a
smaller load carrying capacity compared to the SupplierVehicle. The 
WarehouseVehicle was also designed to have periodic failures (details to follow).

* The capacity of the WarehouseVehicle: *WarehouseVehicleCapacity*

* The frequency of failure: *WarehouseVehicleUptimeBetweenFailures*

The selection weight on the paths from the Warehouse to the Distribution Sites 
were set to dynamically adjust based on the number of Beneficiaries waiting for
Relief Supplies.

### Distribution Sites

Two distribution sites are included in the simulation and are modelled as 
combiners which match Relief Supplies to Beneficiaries.

The following variable was created to facilitate design of experiments:

* The processing time needed to provide a ReliefSupply item to a ReliefBeneficiary: *DistSiteProcessingTime*

Typically this is a fast process, especially with pre-packed relief supplies as
would be expected in the scenario being simulated. A Random Exponential 
with a mean of 1 minute was used to process individuals through the handover of 
supplies from aid worker to beneficiary.

### Relief Beneficiaries

The Beneficiaries represent entities who are affected by the disaster and 
require ReliefSupplies. Two separate sources were included in the model to 
simulate separate points from which beneficiaries would origination for each 
distribution site.

The following variables were created to facilitate design of experiments:

* The interarrival time of the ReliefBeneficiaries: *ReliefBeneficiaryInterarrivalTime*

* The maximum number of ReliefBeneficiaries from a given source: *DisasterVictimMaximumArrivals*

Note that the inter-arrival times of beneficiaries was modelled as a Random 
Exponential event (inter arrival was set at every 2.5 minutes). 

The rational for this is given as follows:

* Field experience of one of the researchers where the process of registrations 
and verification of people has been clocked with an average of 2-3 minutes per 
beneficiary.

* The exponential function is widely used to model arrivals that are completely 
random. As this simulation is concerned with rapid onset emergencies, the 
arrival of beneficiaries will be completely random (contrasted to distributions 
that are more organized ahead of time with the affected communities after weeks 
or months following the disaster).

* Research from studies looking at arrival times at emergency departments 
reference the highly random nature of arrivals. Many use forms of exponential 
family of distributions (such as Weibull)

* Note that this simulation assumes that the source of arrivals are unlimited 
(given the reference to rapid onset and immediate response). Any queue of 
beneficiaries to be modelled, was done so in an infinite manner.

* The exponential nature makes intuitive sense, given that we would expect the 
number of arrivals to taper off as time progresses (people will come to a 
distribution site, but as time passes - people know that the distributions are 
likely closed or the "noise" about a distribution happening in a certain 
location will likely taper off).


### Disaster

The "Disaster" is set to begin at a designated time into the simulation. This 
gives some time to prestage relief supplies through the warehouse (priming the 
pump).

The following variables were created to facilitate design of experiments:

* The amount of time between simulation start and onset of the disaster: *DisasterTimeOffset*


## Measured Outcomes

### Average Beneficiary Time in System

Measured in hours, this response is an important indicator of the performance of
the system. Fewer hours spent in the system are considered to be better outcomes
as people should be looking after their families, rebuilding homes etc rather 
than queuing for relief items.

### Relief Supplies Distributed

How many relief supplies were distributed?

### Unsatisfied Beneficiaries

This was defined as the number of beneficiaries who had yet to receive relief 
items at the conclusion of the simulation. Naturally, we wanted this response to
be zero (0) throughout all simulations.

### Maximum Distribution Site 1 Time Waiting

Measured in hours, this response helped us understand the degree of the worst 
case scenario for distribution site 1 beneficiaries in need of relief supplies.

### Maximum Distribution Site 2 Time Waiting

Measured in hours, this response helped us understand the degree of the worst 
case scenario for distribution site 2 beneficiaries in need of relief supplies.

## Design of Experiments

A variety of experiments were performed in the context of the simulation. The 
following table lists the baseline values for the control variables. Each of the
experiments modifies a single variable and recorded an outcome of interest, as 
described in the following subsections.

```{r, echo=FALSE, tab.cap="Simulation Control Variables"}
dfDefaults <- data.frame(Name=c(
                                "DisasterVictimMaximumArrivals (per Dist Site)",
                                "WarehouseTravelTimeToDistributionSite1",
                                "SupplierMaximumArrivals",
                                "SupplierReliefSuppliesInterarrivalTime",
                                "WarehouseTravelTimeToDistributionSite2",
                                "DisasterTimeOffset",
                                "WarehouseVehicleCapacity",
                                "SupplierVehicleCapacity",
                                "ReliefBeneficiaryInterarrivalTime",
                                "WarehouseVehicleUptimeBetweenFailures",
                                "DistSiteProcessingTime",
                                "WarehouseStorageCapacity",
                                "SupplierTravelTimeToWarehouse",
                                "Entry2DistSitePath"
                                ), 
                         Value=c(
                           10000,
                           "Random.Normal(6,0.5)",
                           "Infinity",
                           "Random.Normal(1, 0.1)",
                           "Random.Normal(6,0.5)",
                           3,
                           1000,
                           5000,
                           "Random.Exponential(2.5)",
                           "Random.Exponential(100)",
                           "Random.Exponential(1)",
                           "Infinity",
                           "Random.Normal(6,0.5)",
                           "TimePathToDistSite2"),
                         Units=c(
                           "Beneficiaries",
                           "Hours",
                           "Relief Supply Items",
                           "Minutes",
                           "Hours",
                           "Days",
                           "Relief Supply Items",
                           "Relief Supply Items",
                           "Minutes",
                           "Hours",
                           "Minutes",
                           "Relief Supply Items",
                           "Hours",
                           "Path Name"
                         ))
rownames(dfDefaults) <- NULL

kable(dfDefaults[order(dfDefaults$Name), ], row.names=FALSE)
```


### Supplier: Partner vs Ad Hoc

How does a dedicated supplier whose production time is lower (more units of 
production are committed to our orders, which can be thought of as being a more 
reliable supplier), versus an ad hoc supplier whose production time is longer, 
affect the measured outcomes?

The following table shows key parameters (control variables) for this scenario:

```{r, echo=FALSE}
dfControls <- data.frame(
  Name=c("Adhoc Supplier", 
         "Midlevel Supplier",
         "Partner Supplier (Baseline)"),
  ReliefSuppliesInterarrivalTime=c("Random.Normal(4, 0.1)",
                                   "Random.Normal(2.5, 0.1)",
                                   "Random.Normal(1, 0.1)"))
kable(dfControls)
```

### Distribution Site: Level of Access

How does the level of access to distribution sites, modelled as travel time 
between the warehouse and distribution site 1, affect outcomes?

```{r, echo=FALSE}
dfControls <- data.frame(
  Name=c("Partner Supplier (Baseline)", 
         "Midlevel Access to Dist Site 1", 
         "Quick Access to Dist Site 1"),
 WarehouseTravelTimeToDistSite1=c("Random.Normal(6,0.5)",
                                   "Random.Normal(4,0.5)",
                                   "Random.Normal(2,0.5)")

)
kable(dfControls)
```

### Warehouse Vehicle: Propensity for Failure

How does the likelihood of the warehouse delivery vehicle failing, affect 
outcomes?

```{r, echo=FALSE}
dfControls <- data.frame(
  Name=c("Partner Supplier (Baseline)", 
         "Warehouse Vehicle Midlevel Failures", 
         "Warehouse Vehicle No Failures"),
  WarehouseVehicleUptimeBetweenFailures=c("Random.Exponential(100)",
                                   "Random.Exponential(50)",
                                   "Infinity")

)
kable(dfControls)
```

### Distribution Site Closure

How does a distribution site closure affect outcomes?

```{r, echo=FALSE}
dfControls <- data.frame(
  Name=c("Partner Supplier (Baseline)", 
         "Distribution Site 2 Closed"),
  Entry2DistSitePath=c("TimePathToDistSite2",
                                          "TimePathToDistSite1")

)
kable(dfControls)
```

# Results

```{r, echo=FALSE}
# Load the results from the Simio CSV
dfResults <- read.table("../project/DisasterReliefSim-Simio_DisasterReliefModel_SupplierExperiment_ResultsSummary.csv",
                        sep=",", header=TRUE, stringsAsFactors=FALSE)

dfResponseData <- read.table("../project/DisasterReliefSim-Simio_DisasterReliefModel_Experiment_ResponseSummary.csv",
                        sep=",", header=TRUE, stringsAsFactors=FALSE)

#summary(dfResults)

# Add on our Control Variable values (haven't found how to export these from Simio yet)
dfResults$SupplierReliefSuppliesInterarrivalTimeMean <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 1
dfResults[dfResults$Scenario == "Midlevel Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 2.5
dfResults[dfResults$Scenario == "Adhoc Supplier", ]$SupplierReliefSuppliesInterarrivalTimeMean <- 4

dfResults$WarehouseTravelTimeToDistributionSite1 <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$WarehouseTravelTimeToDistributionSite1 <- 6
dfResults[dfResults$Scenario == "Midlevel Access to Dist Site 1", ]$WarehouseTravelTimeToDistributionSite1 <- 4
dfResults[dfResults$Scenario == "Quick Access to Dist Site 1", ]$WarehouseTravelTimeToDistributionSite1 <- 2

dfResults$WarehouseVehicleUptimeBetweenFailures <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$WarehouseVehicleUptimeBetweenFailures <- 100
dfResults[dfResults$Scenario == "Warehouse Vehicle Midlevel Failures", ]$WarehouseVehicleUptimeBetweenFailures <- 50
dfResults[dfResults$Scenario == "Warehouse Vehicle No Failures", ]$WarehouseVehicleUptimeBetweenFailures <- 1000


dfResults$DistributionSite2Closed <- NA
dfResults[dfResults$Scenario == "Partner Supplier", ]$DistributionSite2Closed <- 0
dfResults[dfResults$Scenario == "Distribution Site 2 Closed", ]$DistributionSite2Closed <- 1
#dfResults$DistributionSite2Closed <- as.factor(dfResults$DistributionSite2Closed )

```

## Supplier: Partner vs Ad Hoc

```{r, echo=FALSE, fig.height=3}
# Select only the scenarios that are related to this experiment
vSupplierScenarios <- c("Partner Supplier", "Adhoc Supplier", "Midlevel Supplier")
dfExp <- reliefBenAvgForScenarios(vSupplierScenarios, dfResults)

g1 <- ggplot(dfExp) + 
  geom_line(aes(x=SupplierReliefSuppliesInterarrivalTimeMean, y=Average)) +
  labs(x="Supplier Relief Supplies Interarrival Time Mean (minutes)",
       y="Avg Beneficiary Time In System (hours)") +
  myTheme
g1
```

These simulated results are consistent with what we would expect in the real 
world. The investment in establishing strategic agreements with suppliers ahead 
of time (i.e. forming Strategic Partnerships) were modelled with a higher 
"average reliability" in providing the aid agency with supplies (i.e. a faster 
average inter-arrival of supplies). These forms of partnerships clearly had a 
positive impacted in reducing the average time in system for beneficiaries 
further down the chain (at distribution sites). 

The worst performer was the Adhoc supplier relationship, wherein the average 
beneficiary time can be seen to grow to over 250 hours (inconsistent supplies 
meaning that many beneficiaries were kept waiting within the system until the 
pipeline was sufficient to feed the supply needs at the distribution sites). 
Note that the model did not 'open' or 'close' distributions based on set 
operating hours, and as such waiting time accumulated on a 24 hour basis. 
Perhaps this is slightly artificial, however it does indicate the impact of 
unreliable partners further up the supply chain.

```{r, echo=FALSE, fig.height=3}
# Time in system for supply items
dfExp1.2 <- supplyAvgForScenarios(vSupplierScenarios, dfResults)

g1.2 <- ggplot(dfExp1.2) + 
  geom_point(aes(x=Scenario, y=Average)) +
  labs(x="",
       y="Avg Time In System For Supplies (hrs)") +
  myTheme
g1.2
```

The importance of pre-establishing strategic supplier relationships is further 
corroborated in the simulation by looking at the length of time supplies remain
within the system. The ideal scenario would be to have a 'just-in’ time supply
chain thereby minimizing warehouse costs while fully meeting beneficiary needs.
This is especially true for perishable relief items such as food or medicines.
As can be seen in the figure above, the more established a relationship with the 
supplier, the less time it took for supplies to remain within the system.

## Restricted Access to Distribution Site from Warehouse

```{r, echo=FALSE, fig.height=3}
vWarehouseScenarios <- c("Partner Supplier", 
                "Midlevel Access to Dist Site 1", 
                "Quick Access to Dist Site 1")

# Select only the scenarios that are related to this experiment
dfExp <- reliefBenAvgForScenarios(vWarehouseScenarios, dfResults)

#dfExp
g2 <- ggplot(dfExp) + 
  geom_line(aes(x=WarehouseTravelTimeToDistributionSite1, y=Average)) +
  labs(x="Warehouse Travel Time To Distribution Site 1 (hours)",
       y="Avg Beneficiary Time In System (hours)") +
  ylim(0, 10) +
  myTheme
g2
```

An average wait time by beneficiaries of 5 hours is noted when access to a 
distribution site is easier (modelled as quicker time to gain access to the 
distribution site). Interestingly we see that the model suggests that an 
optimal time (in terms of reduced beneficiary wait time at the distribution 
points) to reach/gain access to the distribution site is approximately 4 
hours, after which the average wait time for beneficiaries begins to increase 
again.

It is suggested that the simulation is identifying a phenomenon associated with 
the arrival and queuing times of beneficiaries at a distribution site in 
conjunction with the capacity to serve the queue based on incoming supplies. One
hypothesis is that at  the 4 hour mark, sufficient numbers of beneficiaries have
accumulated to effectively service people more quickly with all available 
supplies arriving on the truck. Quicker arrival times of trucks to the sites may
imply that few people have accumulated in the queues. Moreover longer access 
times of the vehicles may mean that the buffer queue of beneficiaries is not 
being cleared in the most optimal manner (Note that the capacity of the trucks 
from the warehouse to the distribution sites are limited to carrying 1000 units 
per trip).

## Warehouse Vehicle Failures

The warehouse vehicle failures yielded interesting results. It appears the 
difference between a 100 hour exponential mean and no failures is negligible. 
Strangely, the outcome of average beneficiaries time in system reduces for the 
mean failure rate of 50 uptime hours between failures. The confidence interval 
for these values reveals that an overlap suggesting an insignificant difference.

```{r, echo=FALSE, fig.height=3}
vWarehouseVehicleScenarios <- c("Partner Supplier", 
                "Warehouse Vehicle Midlevel Failures", 
                "Warehouse Vehicle No Failures")

# Select only the scenarios that are related to this experiment
dfExp <- reliefBenAvgForScenarios(vWarehouseVehicleScenarios, dfResults)

#dfExp
g2 <- ggplot(dfExp) + 
  geom_line(aes(x=WarehouseVehicleUptimeBetweenFailures, y=Average)) +
  labs(x="Warehouse Vehicle Update Between Failures (hours)",
       y="Avg Beneficiary Time In System (hours)") +
    ylim(0, 10) +
  myTheme
g2
```

## Distribution Site Closure

The distribution site closure did not have much of an effect on the primary outcome, average beneficiaries time in system. This requires further investigation but could indicate the travel time for the extra beneficiaries was too low. Additionally, it seems the capacity of the relief distribution system (as modelled) has sufficient head room to take on the additional demand at a single distribution site.

```{r, echo=FALSE, fig.height=3}
vDistSiteScenarios <- c("Partner Supplier", 
                "Distribution Site 2 Closed")

# Select only the scenarios that are related to this experiment
dfExp <- reliefBenAvgForScenarios(vDistSiteScenarios, dfResults)

#dfExp
g2 <- ggplot(dfExp) + 
  geom_line(aes(x=DistributionSite2Closed, y=Average)) +
  labs(x="Distribution Site 2 Closed (boolean)",
       y="Avg Beneficiary Time In System (hours)") +
  ylim(0, 10) +
  myTheme
g2
```

While the average beneficiary time in the system when distribution site 2 was 
closed, did not substantially differ from other experimental scenarios (with the 
exception of the noted differences when supplier was not equal to a Partner 
supplier), it was noted that the length of time for supplies to move through the
system was slightly quicker when site 2 was closed. The range is seen to be lower 
under this scenario:

\begin{center}
\includegraphics{../images/SimioReliefSuppliesTimeInSystem.png}
\end{center}
\begin{center}
\text{All Scenarios - Relief Supplies Time In System}
\end{center}

Note that the average travel time between warehouses and distribution sites was 
the same (6 hours, with standard deviation of 30 minutes). Further, there is 
only one truck servicing both distribution sites. It is unclear as to why 
supplies appeared to move more quickly through the system in this scenario. One 
hypothesis is that when site 2 was closed, the queue of beneficiaries at site 1 
filled up quicker and the numbers where therefore greater when distributing 
supplies following the arrival of the truck. In essence, the simulation may 
indicate that closing site 2, presented an opportunity to serve more people in a
single truck-load of relief items thereby reducing the time for supplies to be 
in the system. The average wait time of beneficiaries would not have differed as
the number of people entering the simulation did not change.

# Suggestions for future work 

The current study provides a good indication of in-country supply chain dynamics 
down to the last mile of distributions to beneficiaries. The underlying 
assumptions associated with production frequencies based on supplier type, the 
expected inter-delivery times of supplies, and the arrival behaviour of 
beneficiaries were estimated using domain knowledge experience and some 
limited field data. However much of this needs to be supplemented with  
experiments and production time observations at a broader scale. 

It is also acknowledged that the context of the disaster will impact results. 
For example, disasters significantly affecting transportation infrastructure, 
would necessitate modelling different modes of delivery (such as aerial) instead
of land-based vehicles. Future work may also look to adding an aspect of 
disaster intensity and disaster frequency as additions to the simulation.

Further, adding a financial analysis component to the the models to provide a 
cost quantification metrics would be a natural extension. For example to 
generate cost-benefit analysis of the different modes of reliable suppliers. 
This could be extended to include the opportunity costs of wait times for 
beneficiaries.

Quantifying supplies would help in finding optimal solutions in terms of the 
amount of stock to preposition, versus time to deliver relief (efficiency) and
also to compare against wastage (spoiled supplies).

Finally, future work may wish to consider growing the entities to encompass more
complexity associated with local supply chain systems. For example, modelling a 
greater number of warehouse centres (possibly modelling the movement from an 
in-country central warehouse to district level warehouses that hold supplies 
prior to running pre-planned distributions). The notion of security and risk may
consequently be built into such extended models, as the security risks at 
district level warehouses are typically higher than those at central warehouses.

# Summary

The simulation model developed in this study is focused on an in-country
disaster relief scenario. The model is capable of more experiments than 
were executed given the variables and capabilities of Simio, but those 
experiments which were performed show that supplier capability and commitment
can have a significant impact on beneficiary outcomes. Additionally, depending
on supply chain capability and other factors, more distribution sites may actually 
degrade overall beneficiairy outcomes. Opening an additionaly distribution site 
should be considered carefully against the impact to existing sites.

All materials related to this study including the actual Simio simulation model 
are shared on GitHub under our DisasterReliefSim repository `r citep(bibGitHub)`.

# References

```{r, results='asis', echo=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

# Appendix: Response Data

```{r, echo=FALSE, tab.cap="Supplier: Partner vs Ad Hoc Responses"}
df <- dataForScenarios(vSupplierScenarios, dfResponseData)
kable(df)
```

```{r, echo=FALSE, tab.cap="Warehouse Access to Distribution Site Responses"}
df <- dataForScenarios(vWarehouseScenarios, dfResponseData)
kable(df)
```

```{r, echo=FALSE, tab.cap="Warehouse Access to Distribution Site Responses"}
df <- dataForScenarios(vWarehouseScenarios, dfResponseData)
kable(df)
```

\begin{center}
\includegraphics{../images/SimioAvgBenTimeInSys.png}
\end{center}
\begin{center}
\text{All Scenarios - Average Beneficiaries Time In System}
\end{center}

\begin{center}
\includegraphics{../images/SimioScreenshot3d.png}
\end{center}
\begin{center}
\text{Simio Simulation Model - 3D}
\end{center}
