---
title: "Final Project Proposal"
author: "Daniel Dittenhafer & Jashan Narhan"
date: "March 9, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 604: Simulation & Modeling"
---
```{r, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(RefManageR)

cleanbib()

cite_options(style="markdown")

# bibBirthData <- bibentry(bibtype="Misc",
#                          author=person(family="HHS"),
#                          publisher="United States Department of Health and Human Services (US DHHS), Centers for Disease Control and Prevention (CDC), National Center for Health Statistics (NCHS), Division of Vital Statistics",
#                          title="Natality public-use data 2007-2014 on CDC WONDER Online Database",
#                          year=2016,
#                          month="February",
#                          url="http://wonder.cdc.gov/natality-current.html")


bibBastianEtAl <- bibentry(bibtype="Article",
                 title="Multi-criteria logistics modeling for military humanitarian assistance and disaster relief aerial delivery operations",
                 author=personList(person(family="Bastian", given="Nathaniel"),
                                   person(family="Griffin", given="Paul"),
                                   person(family="Spero", given="Eric"),
                                   person(family="Fulton", given="Lawrence")),
                 publisher="Springer Berlin Heidelberg",
                 journal="Optimization Letters",
                 year=2015,
                 doi="10.1007/s11590-015-0888-1")

```
# Problem Statement & Research Questions

TBD

# Statement of Objectives

TBD

# Data Set(s)

TBD



# Recent Journal Papers

## Multi-criteria logistic modeling for miliary humanitarian assistance...

This paper describes a model complete with objective function an array of constraints for aiding in decision support for humanitarian assistance and disaster relief (HA/DR) efforts. Specifically, the model aims to determine the optimal the aerial supply chain network for HA/DR given target goals for response time, supply chain cost and unmet demand in the disaster zone (`r citep(bibBastianEtAl)`).

## Article 2

TBD

## Evaluation

TBD

# References

```{r, results='asis', echo=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

